Women in Film

### 2017-08-22 draft_post.Rmd
Some edits, based on feedback from copy editor

How are women portrayed in film? It’s complicated; that question might make you think of a male protagonist’s romantic interest, a damsel in distress who needs to be rescued, or Wonder Woman.

One way to quantitatively measure how women are portrayed in film is to look at scene directions in film scripts. Scene directions are the part of a script that are not dialogue, describing what’s happening on screen and telling actors how to move and how to deliver their lines.

**In this analysis, we used text mining to identify words that are paired with the pronouns “he” and “she” in scene directions for about 2,000 film scripts for movies released from the 1930s to today.** We then analyzed which words are more likely to come after one or the other.

@@

There are dramatic differences in the kinds of words we see here. In movies, women are more likely to snuggle, giggle, squeal, and sob while men are more likely to strap things on, gallop, shoot, howl, and kill. We see evidence here of the tropes and storytelling conventions that propel so many films, and how men, women, and their relationships are often portrayed.

Now let’s look at the top 800 words, including those that are used about evenly after “he” and “she”.

@@

Hollywood can be a tough place to work as a creative woman, either writing, directing, or producing films. We used our text mining approach to explore the impact that the gender of a writer has on how characters are portrayed. Using IMDB biographies, pictures, and names, we manually classified the genders of writers for our film script dataset.

English has two singular third person pronouns most often used for people, “he” and “she”. In this analysis, for both the text data and the identification of gender for film writers, we have chosen to identify men and women with the pronouns “he” and “she”. Using this type of classification, anyone (writer or character) that uses the pronoun “she” would be classified as a woman.

With these data, we can examine how male and female writers portray women in film. We can dig deeper into the kind of analysis we showed above and see how the results depend on who is writing a character. Do women writers portray women differently? What are the words that both male and female writers use about equally when crafting characters who are women, or men? What are the words that they use differently? How would the overall cultural representation of women be different if there were more women writing the movies we watch? For this part of the analysis, we will narrow the focus to the most commonly used 400 words.

@@

In our dataset, about 15% of film writers were women and about 85% were men. If women were more equitably represented in the ranks of Hollywood writers, our analysis indicates we would see fewer female characters responding, kissing, and crying and more women spying, writing, and finding things.

There are some words that both male and female writers are likely to use when directing a character of a certain gender.. We see that women in film are more likely, compared to male characters, to gasp, hurry, smile, hesitate, and stir (mostly while cooking), regardless of whether the writer is a man or a woman. Men in film are more likely, compared to women, to smash things, draw their weapons, grin, wink, point, talk *and* speak, when being written by both male and female screenwriters.

We can also see remarkable differences in how male and female writers portray characters who are men and women. When describing the opposite gender, both men and women use some overtly romantic and sexual words like “kiss”, and “stroke”, as well as more subtle words like “respond” and “embrace”. When writing for the opposite gender, writers also use words that imply, well, less happy outcomes, such as “cry”, “glare”, and “died”.

When characters are penned by an author of their same gender, they are more likely to wince, jerk, fumble, and perhaps most remarkably, write on screen. There are verbs here that are active but less explicitly testosterone-fueled than when writers create male characters; when writers create characters the same gender as themselves, these characters spy, scan, sprint, and climb. These characters also search, remember, and realize. A woman in a movie doing such things is a far cry from the giggling, squealing, blushing character we met at the beginning of this essay.

> The code used in analysis is [publicly available on GitHub](https://github.com/juliasilge/women-in-film). The data set for this analysis included 1,966 scripts for films released between 1929 and 2015; most are from 1990 and after. Each script was processed to extract only the set directions, excluding dialogue from this analysis. We then identified all [bigrams in these scripts that had either “he” or “she” as the first word in the bigram](https://juliasilge.com/blog/gender-pronouns). Then, we calculated a log odds ratio to find words that exhibit the biggest differences between relative use for “she” and “he”. We removed stop words and did some other minimal text cleaning to maintain meaningful results. We calculated the overall log odds ratio for the 800 most commonly used words, and then log odds ratios for only scripts with male writers and female writers for the 400 most commonly used words. Scripts often have more than one writer and could be counted in both categories. To learn more about text mining analyses like this one and how to perform them, [check out Julia’s book](http://tidytextmining.com).

>>>>

In April 2016, we [broke down](https://pudding.cool/2017/03/film-dialogue/) film dialogue by gender. The essay presented an imbalance in which men delivered more lines than women across 2,000 screenplays. But quantity of lines is only part of the story. What characters do matters, too.

Gender tropes (e.g., [women are pretty/men act, men don’t cry](http://tvtropes.org/pmwiki/pmwiki.php/Main/MenAreStrongWomenArePretty)) are just as important as dialogue in understanding how men and women are portrayed on-screen. These stereotypes result from many components, including casting, acting, directing, etc.

The film script, arguably, is ground zero—the source material by which everyone is influenced. And in film scripts, there’s dialogue and screen direction. For example, let’s take this iconic scene from Titanic:

> compare dialogue/set directions here

The curious data here is less what Rose says (“I’m flying”) and more what the screen direction prescribes (“she smiles dreamily,” “he pushes against her”). In the following analysis, we go deep on screen direction to understand gender tropes. **We examined 2,000 scripts and broke down every screen direction mapped to the pronouns “he” and “she.”**

@@

These are the most extreme examples. There is a high likelihood that women will snuggle, giggle, squeal, and sob, relative to men. Conversely, men are more likely to strap, gallop, shoot, howl, and kill.

Let’s now examine the 800 most commonly used pronoun pairs in screen direction.

@@

Next, let’s examine how the writer’s gender affects characters’ behavior. Do women writers use different language for women roles? What are the words that both male and female writers use about equally when describing characters? Would results change dramatically if there were more women writers? First, we will narrow the data set to the most commonly used 400 words.

@@

There are some directions where the writer’s gender makes no difference. Relative to men, women gasp, hurry, smile, hesitate, and stir (mostly while cooking), regardless of whether the writer is a man or a woman. Men are consistently more likely to smash things, draw their weapons, grin, wink, point, talk, and speak.

When describing the opposite gender, both men and women use some overtly romantic and sexual words, such as “kiss” and “stroke,” as well as more subtle words including “respond” and “embrace.”

But there are differences. In our data set, 15% of film writers were women; 85% were men. Should Hollywood reach gender parity, we’d expect fewer women characters to respond, kiss, and cry. The increase in female writers would also mean women would be more likely to spy, find things, and, perhaps most remarkably, write on-screen.

> The code used in analysis is [publicly available on GitHub](https://github.com/juliasilge/women-in-film). The data set for this analysis included 1,966 scripts for films released between 1929 and 2015; most are from 1990 and after. Each script was processed to extract only the set directions, excluding dialogue from this analysis. We then identified all [bigrams in these scripts that had either “he” or “she” as the first word in the bigram](https://juliasilge.com/blog/gender-pronouns).

> Then, we calculated a log odds ratio to find words that exhibit the biggest differences between relative use for “she” and “he”. We removed stop words and did some other minimal text cleaning to maintain meaningful results. We calculated the overall log odds ratio for the 800 most commonly used words, and then log odds ratios for only scripts with male writers and female writers for the 400 most commonly used words. Scripts often have more than one writer and could be counted in both categories. To learn more about text mining analyses like this one and how to perform them, [check out Julia’s book](http://tidytextmining.com).

> Writers’ gender was determined via IMDB biographies, pictures, and names.

> English has two singular third-person pronouns most often used for people, “he” and “she.” In this analysis, for both the text data and the identification of gender for film writers, we have chosen to identify men and women with the pronouns “he” and “she.” Using this type of classification, any writer or character associated with the pronoun “she” is classified as a woman.
